Assessing spatial correlations in pulsar timing arrays with the noise-marginalized optimal statistic

### 2017-08-04 manuscript/manuscript.tex
edits to sky scramble section

We used these three techniques to compute the optimal statistic for simulated data sets. 
We constructed simulated data sets using the same procedure as in \citet{tlb+2017}. 
The simulated PTAs consisted of 10 MSPs with \sv{Steve fill in here...}, 
and the injected GWB amplitude was $\Agw = 5\times10^{-15}$. 
The optimal statistic was computed using the \sv{fill in...} in PAL2 \sv{add citation}.

Figure~\ref{fig:os_dataset_sample} shows the fixed-noise and noise-marginalized 
optimal statistic for a single realization of the stochastic background. 
We also include the distribution of $\Agw^2$ found using the results of a Bayesian analysis 
for a common red process. For this particular realization of the stochastic background, 
the values of $\hat{A}^2$ from the noise-marginalized and Bayesian analyses are in good agreement 
with each other and the injected GW stochastic background, 
but the value from the fixed-noise analysis is significantly lower. 
The noise-marginalized analysis also gives a larger mean SNR compared to the fixed-noise analysis.
\begin{figure}[ht]
	\includegraphics[width=0.9\columnwidth]{plots/os_dataset50.pdf}
	\caption{Optimal statistic and SNR for a simulated dataset 
			containing an injected GW background with $\Agw = 5\times10^{-15}$. 
			We compare the values found using a fixed-noise analysis (dashed blue line) to the 
			mean optimal statistic and mean SNR found by marginalizing over 
			the individual pulsars' red noise parameters (solid orange line). 
			We also include the distribution of $\Agw^2$ found using the results of a Bayesian analysis 
			for a common red process (solid green line). 
			The dotted vertical line indicates the injected value, $\hat{A}^2 = 2.5 \times 10^{-29}$. 
			For this particular realization of the stochastic background, the fixed-noise analysis 
			gives $\hat{A}^2 = 1.9 \times 10^{-30}$ with $\mathrm{SNR} = 0.60$, 
			while the noise-marginalized analysis gives $\hat{A}^2 = 1.9\times10^{-29}$ with $\mathrm{SNR} = 1.1$. 
			For comparison, the Bayesian analysis gives a mean value of $\Agw^2 = 4.5\times10^{-29}$.}
	\label{fig:os_dataset_sample}
\end{figure}

In Fig.~\ref{fig:os_datasetstats} we compare the fixed-noise values of $\hat{A}^2$ 
to the mean $\hat{A}^2$ for 425 different realizations of the stochastic background. 
The fixed-noise analysis tends to underestimate the strength of the stochastic background, 
whereas the noise-marginalized analysis is better able to recover the injected value. 
The mean value of the optimal statistic from the fixed-noise analysis, 
averaged over realizations of the stochastic background, is $\hat{A}^2 = 8.9 \times 10^{-30}$, 
while the mean value from the noise-marginalized analysis is $\hat{A}^2 = 2.5 \times 10^{-29}$. 
The noise-marginalized analysis also gives a higher average mean SNR of 1.7, 
compared to an average SNR of 1.0 using the fixed-noise analysis.
\begin{figure}[ht]
	\includegraphics[width=0.9\columnwidth]{plots/os_datasetstats_5e-15.pdf}
	\caption{Optimal statistic and SNR for 425 simulated datasets 
			containing an injected GW background with $\Agw = 5\times10^{-15}$. 
			We compare the values found using a fixed-noise analysis (blue) to the 
			mean optimal statistic and mean SNR found by marginalizing over 
			the individual pulsars' red noise parameters. The dashed vertical line 
			indicates the injected value, $\hat{A}^2 = 2.5 \times 10^{-29}$. 
			The value of $\hat{A}^2$ from the fixed-noise analysis, 
			averaged over realizations of the stochastic background, is $\hat{A}^2 = 8.9 \times 10^{-30}$, 
			while the noise-marginalized analysis gives a mean value of $\hat{A}^2 = 2.5 \times 10^{-29}$. 
			Furthermore, the fixed-noise analysis gives an average SNR of 1.0, while the 
			noise-marginalized analysis gives an average mean SNR of 1.7.}
	\label{fig:os_datasetstats}
\end{figure}

\section{Compare Significance with Monopole and Dipole}
\label{sec:spatial}

@@

Figure~\ref{fig:os_ORF} shows the noise-marginalized mean value of $\hat{A}^2$ and the mean SNR 
computed assuming monopole, dipole, and quadrupole spatial correlations 
for 425 simulated data sets. Using a monopole or dipole ORF 
gives a lower value for the mean optimal statistic and mean SNR compared to the 
Hellings-Downs ORF. 
However, there is significant overlap between the distributions for the mean SNR using 
monopole, dipole, and quadrupole ORFs, indicating that it is not easy to differentiate 
between those three spatial correlations for an injected GW background of this amplitude. 
We find a noise-marginalized mean SNR above 1.0 in 68\% of our simulated data sets 
using the quadrupole ORF, and in 45\% and 24\% of our simulated data sets 
using the monopole and dipole ORF, respectively.
\begin{figure}[ht]
	\includegraphics[width=0.9\columnwidth]{plots/os_datasetstats_5e-15_ORF.pdf}
	\caption{Noise-marginalized mean optimal statistic and mean SNR for 425 simulated datasets 
			containing an injected GW background with $\Agw = 5\times10^{-15}$. 
			We compare the values found using monopole (blue), dipole (orange), and quadrupole (green) 
			spatial correlations. 
			The dashed vertical line indicates the injected value, $\hat{A}^2 = 2.5 \times 10^{-29}$.}
	\label{fig:os_ORF}
\end{figure}

\section{Compare Significance with Sky Scrambles/Phase Shifts}
\label{sec:skyscrambles}

@@

The scrambled sky positions were generated \sv{Steve, explain here...}

Figure~\ref{fig:skyscrambles_dataset_sample} compares the noise-marginalized optimal statistic 
for a particular data set to the distribution of noise-marginalized optimal statistics obtained using sky scrambles. 
For this data set, the noise-marginalized mean optimal statistic is 
$1.9\times10^{-29}$ and the mean SNR is 1.1. 
The distributions of the noise-marginalized optimal statistic and SNR obtained using the scrambled ORFs 
are both centered around zero, 
and for 24 out of the 210 sky scrambles, the SNR is above the true value of $1.1$ ($p=0.11$). 
Figure~\ref{fig:skyscrambles} shows the $p$-values for 210 sky scrambles 
using 425 realizations of the stochastic background. 
For 62\% of simulations the $p$-value is $<0.1$, and for 29\% of simulations the $p$-value is $<0.01$.
\begin{figure}[ht]
	\includegraphics[width=0.9\columnwidth]{plots/skyscrambles_dataset50.pdf}
	\caption{Comparison between the optimal statistic with and without sky scrambles 
			for a simulated dataset containing an injected GW background with $\Agw = 5\times10^{-15}$. 
			The true analysis found an $\mathrm{SNR} = 1.1$, while only 24 of the 210 sky scramble analyses 
			found $\mathrm{SNR} \geq 1.1$ ($p=0.11$).}
	\label{fig:skyscrambles_dataset_sample}
\end{figure}
\begin{figure}[ht]
	\includegraphics[width=0.9\columnwidth]{plots/os_skyscrambles.pdf}
	\caption{Distribution of $p$-values for 425 different realizations of the stochastic background. 
			For each realization, the $p$-value is the fraction of sky scrambled ORFs that gives a 
			noise-marginalized mean SNR greater than the mean SNR using the correct ORF.}
	\label{fig:skyscrambles}
\end{figure}

@@

\acknowledgments
We thank Joe Romano and Xavier Siemens for useful discussions. 
JAE was supported by NASA through Einstein Fellowship Grant PF4-150120. 
SRT was supported by appointment to the NASA Postdoctoral Program 
at the Jet Propulsion Laboratory, which is administered by Oak Ridge Associated Universities 
and the Universities Space Research Association through a contract with NASA. 
KPI and SJV are supported by NSF Physics Frontier Center Grant 1430284.

>>>>

We used these three techniques to compute the optimal statistic for simulated data sets. 
We constructed simulated data sets using the same procedure as in \citet{tlb+2017}. 
The simulated PTAs consisted of 10 MSPs with 
total observation times and noise properties chosen to be 
similar to the \sv{Steve fill in here...} (see Table~\ref{tab:sim}). 
We also injected a stochastic GWB amplitude with $\Agw = 5\times10^{-15}$.
\begin{table}[htb]
	\setlength{\tabcolsep}{5pt}
	\caption{Pulsar parameters used in simulated PTA}
	\begin{center}
	\begin{tabular}{ccccc}
		\hline\hline
    		Pulsar	& $T_\mathrm{obs}$ (yrs) & $\sigma_w$ ($\mu\mathrm{s}$) & $\log_{10}A_\mathrm{red}$ & $\gamma_\mathrm{red}$ \\
		\hline
		J0613$-$0200 & 16.054 & 1.58 & -13.90 & 3.18 \\
		J0751+1807 & 17.606 & 2.60 & -14.14 & 2.58 \\
		J1012+5307 & 16.831 & 1.47 & -13.09 & 1.65 \\
		J1640+2224 & 16.735 & 1.99 & -13.24 & 0.03 \\
		J1643$-$1224 & 17.300 & 1.65 & -18.56 & 4.04 \\
		J1713+0747 & 17.657 & 0.26 & -14.90 & 4.85 \\
		J1744$-$1134 & 17.250 & 0.65 & -13.60 & 2.00 \\
		J1857+0943 & 17.310 & 1.51 & -16.00 & 1.35 \\
		J1909$-$3744 & 9.379 & 0.12 & -13.99 & 2.06 \\
		J2145$-$0750 & 17.161 & 1.19 & -13.87 & 4.02 \\
    		\hline\hline
	\end{tabular}
	\end{center}
	\label{tab:sim}
\end{table}

The optimal statistic was computed using the \sv{fill in...} in PAL2 \sv{add citation}. 
Figure~\ref{fig:os_dataset_sample} shows the fixed-noise and noise-marginalized 
optimal statistic for a single realization of the stochastic background. 
We also include the distribution of $\Agw^2$ found using the results of a Bayesian analysis 
for a common red process. For this particular realization of the stochastic background, 
the values of $\hat{A}^2$ from the noise-marginalized and Bayesian analyses are in good agreement 
with each other and the injected GW stochastic background, 
but the value from the fixed-noise analysis is significantly lower. 
The noise-marginalized analysis also gives a larger mean SNR compared to the fixed-noise analysis.
\begin{figure}[ht]
	\includegraphics[width=0.95\columnwidth]{plots/os_dataset50.pdf}
	\caption{Optimal statistic and SNR for a simulated dataset 
			containing an injected GW background with $\Agw = 5\times10^{-15}$. 
			We compare the values found using a fixed-noise analysis (dashed blue line) to the 
			mean optimal statistic and mean SNR found by marginalizing over 
			the individual pulsars' red noise parameters (solid orange line). 
			We also include the distribution of $\Agw^2$ found using the results of a Bayesian analysis 
			for a common red process (solid green line). 
			The dotted vertical line indicates the injected value, $\hat{A}^2 = 2.5 \times 10^{-29}$. 
			For this particular realization of the stochastic background, the fixed-noise analysis 
			gives $\hat{A}^2 = 1.9 \times 10^{-30}$ with $\mathrm{SNR} = 0.60$, 
			while the noise-marginalized analysis gives $\hat{A}^2 = 1.9\times10^{-29}$ with $\mathrm{SNR} = 1.1$. 
			For comparison, the Bayesian analysis gives a mean value of $\Agw^2 = 4.5\times10^{-29}$.}
	\label{fig:os_dataset_sample}
\end{figure}

In Fig.~\ref{fig:os_datasetstats} we compare the fixed-noise values of $\hat{A}^2$ 
to the mean $\hat{A}^2$ for 425 different realizations of the stochastic background. 
The fixed-noise analysis tends to underestimate the strength of the stochastic background, 
whereas the noise-marginalized analysis is better able to recover the injected value. 
The mean value of the optimal statistic from the fixed-noise analysis, 
averaged over realizations of the stochastic background, is $\hat{A}^2 = 8.9 \times 10^{-30}$, 
while the mean value from the noise-marginalized analysis is $\hat{A}^2 = 2.5 \times 10^{-29}$. 
The noise-marginalized analysis also gives a higher average mean SNR of 1.7, 
compared to an average SNR of 1.0 using the fixed-noise analysis.
\begin{figure}[ht]
	\includegraphics[width=0.95\columnwidth]{plots/os_datasetstats_5e-15.pdf}
	\caption{Optimal statistic and SNR for 425 simulated datasets 
			containing an injected GW background with $\Agw = 5\times10^{-15}$. 
			We compare the values found using a fixed-noise analysis (blue) to the 
			mean optimal statistic and mean SNR found by marginalizing over 
			the individual pulsars' red noise parameters. The dashed vertical line 
			indicates the injected value, $\hat{A}^2 = 2.5 \times 10^{-29}$. 
			The value of $\hat{A}^2$ from the fixed-noise analysis, 
			averaged over realizations of the stochastic background, is $\hat{A}^2 = 8.9 \times 10^{-30}$, 
			while the noise-marginalized analysis gives a mean value of $\hat{A}^2 = 2.5 \times 10^{-29}$. 
			Furthermore, the fixed-noise analysis gives an average SNR of 1.0, while the 
			noise-marginalized analysis gives an average mean SNR of 1.7.}
	\label{fig:os_datasetstats}
\end{figure}

\section{Monopole and Dipole Spatial Correlations}
\label{sec:spatial}

@@

Figure~\ref{fig:os_ORF} shows the noise-marginalized mean value of $\hat{A}^2$ and the mean SNR 
computed assuming monopole, dipole, and quadrupole spatial correlations 
for 425 simulated data sets. Using a monopole or dipole ORF 
gives a lower value for the mean optimal statistic and mean SNR compared to the 
Hellings-Downs ORF. 
However, there is significant overlap between the distributions for the mean SNR using 
monopole, dipole, and quadrupole ORFs, indicating that it is not easy to differentiate 
between those three spatial correlations for an injected GW background of this amplitude. 
We find a noise-marginalized mean SNR above 1.0 in 68\% of our simulated data sets 
using the quadrupole ORF, and in 45\% and 24\% of our simulated data sets 
using the monopole and dipole ORFs, respectively.
\begin{figure}[htb]
	\includegraphics[width=0.95\columnwidth]{plots/os_datasetstats_5e-15_ORF.pdf}
	\caption{Noise-marginalized mean optimal statistic and mean SNR for 425 simulated datasets 
			containing an injected isotropic stochastic GWB with $\Agw = 5\times10^{-15}$. 
			We compare the values found using monopole (blue), dipole (orange), 
			and quadrupole (green) spatial correlations. 
			The dashed vertical line indicates the injected value, $\hat{A}^2 = 2.5 \times 10^{-29}$.}
	\label{fig:os_ORF}
\end{figure}

\section{Sky Scrambles}
\label{sec:skyscrambles}

@@

We used the same scrambled sky positions as were used in \citet{tlb+2017}. 
These positions were chosen so as to minimize the overlap  
between the scrambled ORFs and the true ORF, 
which can be assessed using the ``match statistic'' \citep{cs2016}:
\begin{equation}
	\bar{M} = \frac{\sum_{I,J \neq I} \Gamma_{IJ} \Gamma'_{IJ}}{\sqrt{ \left( \sum_{I, J \neq I} \Gamma_{IJ} \Gamma_{IJ} \right) \left( \sum_{I, J \neq I} \Gamma'_{IJ} \Gamma'_{IJ} \right)}} \,,
\end{equation}
where $I$,$J$ index the pulsar number, and $\Gamma$ and $\Gamma'$ are two different ORFs. 
The 210 sky scrambles used in this paper were generated using a particle swarm optimization (PSO) algorithm 
with the requirement that the scrambled ORFs had $\bar{M} < 0.1$ when compared to the true ORF and to each other.
\sv{Steve, check this section}

Figure~\ref{fig:skyscrambles_dataset_sample} compares the noise-marginalized optimal statistic 
for a particular data set to the distribution of noise-marginalized optimal statistics obtained using sky scrambles. 
For this data set, the noise-marginalized mean optimal statistic is 
$1.9\times10^{-29}$ and the mean SNR is 1.1. 
The distributions of the noise-marginalized optimal statistic and SNR obtained using the scrambled ORFs 
are both centered around zero, 
and for 24 out of the 210 sky scrambles, the SNR is above the true value of $1.1$ ($p=0.11$). 
Figure~\ref{fig:skyscrambles} shows the $p$-values for 210 sky scrambles 
using 425 realizations of the stochastic background. 
For 62\% of simulations the $p$-value is $<0.1$, and for 29\% of simulations the $p$-value is $<0.01$.
\begin{figure}[htb]
	\includegraphics[width=0.95\columnwidth]{plots/skyscrambles_dataset50.pdf}
	\caption{Comparison between the noise-marginalized mean optimal statistic and mean SNR 
			with and without sky scrambles for a simulated dataset 
			containing an injected GW background with $\Agw = 5\times10^{-15}$. 
			Using the true ORF, we found an SNR of 1.1, while only 24 of the 210 analyses using scrambled ORFs 
			found an SNR $\geq 1.1$ ($p=0.11$).}
	\label{fig:skyscrambles_dataset_sample}
\end{figure}
\begin{figure}[htb]
	\includegraphics[width=0.95\columnwidth]{plots/os_skyscrambles.pdf}
	\caption{Distribution of $p$-values for 425 different realizations of the stochastic background. 
			For each realization, the $p$-value is the fraction of sky scrambled ORFs that gives a 
			noise-marginalized mean SNR greater than the mean SNR using the correct ORF.}
	\label{fig:skyscrambles}
\end{figure}

@@

The noise-marginalized optimal statistic is a computationally inexpensive 
method for assessing the significance of an isotropic stochastic GWB in 
PTA data.

\acknowledgments
We thank Joe Romano and Xavier Siemens for useful discussions. 
KPI and SJV are supported by NSF Physics Frontier Center Grant 1430284.
JAE was supported by NASA through Einstein Fellowship Grant PF4-150120. 
SRT was supported by appointment to the NASA Postdoctoral Program 
at the Jet Propulsion Laboratory, which is administered by Oak Ridge Associated Universities 
and the Universities Space Research Association through a contract with NASA.
